\newpage
\section{Acknowledgements}

I am grateful to the members of the Loco project at
\MyUrl{http://www.inria.fr/Unites/ROCQUENCOURT-eng.html} {INRIA
  Rocquencourt} for their encouragement. Their involvement in this
work led to useful feedback and exchange (1995).

Many thanks to all those people at \MyUrl{http://www.gnu.org}{GNU} who helped
me to finalize the GNU Prolog project (1999).

I would like to thank everybody who tested preliminary releases and
helped me to put the finishing touches to this system (1999).

I would particularly like to thank
\MyUrl{http://www.sju.edu/\Tilde{jhodgson}}{Jonathan Hodgson} for the time and
effort he put into the proofreading of this manual. His suggestions, both
regarding ISO technical aspects as well as the language in which it was
expressed, proved invaluable (1999-2017).

Thanks to Richard A. O'Keefe for his advice regarding the implementation of
some Prolog built-in predicates and for suggesting me the in-place
installation feature (1999).

The on-line HTML version of this document was created using
\MyUrl{http://pauillac.inria.fr/\Tilde{maranget}/hevea/}{\hevea} developed by
Luc Maranget who kindly devoted so much
of his time extending the capabilities of {\hevea} in order to handle such a
sizeable manual (2000).

Jean-Christophe Aude kindly improved the visual aspect of both the
illustrations and the GNU Prolog web pages (2000).


Many thanks to the following contributors:

\begin{itemize}

\item \MyEMail{clive@laluna.demon.co.uk}{Clive Cox} and
\MyUrl{http://www.rano.org/}{Edmund Grimley Evans} for their port to
ix86/SCO (1999).

\item \MyUrl{http://www.speech.sri.com/people/stolcke/}{Andreas Stolcke} for
his port to ix86/Solaris (1999).

\item \MyEMail{ax@apax.net}{Alexander Diemand} for his initial port to
alpha/linux (2000).

\item \MyEMail{nollinge@ens-lyon.fr}{Nicolas Ollinger} to for his port to
ix86/FreeBSD (2000).

\item \MyEMail{brook@nmsu.edu}{Brook Milligan} for his port to
ix86/NetBSD and for general configuration improvements (2000).

\item \MyEMail{spratt@alum.mit.edu}{Lindsey Spratt} for his port to
powerpc/Darwin (MacOS X) (2002).

\item \MyEMail{gbeauchesne@mandrakesoft.com}{Gwenol\'{e} Beauchesne} for his port to
x86\_64/Linux (2002).

\item \MyEMail{jtb@netbsd.org}{Jason Beegan} for his port to
sparc/NetBSD and to powerpc/NetBSD (2003).

\item \MyEMail{csrabak@ig.com.br}{Cesar Rabak} for his initial port to
ix86/MinGW (2004).

\item \MyEMail{Scott@coral8.com} {Scott L. Burson} for his port to x86\_64/Solaris (2007).

\item \MyEMail{dholland@netbsd.org} {David Holland} for his port to x86\_64/BSD systems (2008).

\item \MyEMail{gclient.gaap@gmail.com} {Ozaki Kiichi} for his port to x86\_64/Darwin (Mac OS X) (2012).

\item \MyEMail{jasper@simulistics.com} {Jasper Taylor} for his port to x86\_64/MinGW64 (2011).

\item \MyEMail{jasper@simulistics.com} {Jasper Taylor} for his ports to arm32/Linux and arm64/Linux (2021).

\end{itemize}

Many thanks to Paulo Moura for his continuous help (in particular
about Darwin ports), for his ISO Prolog unit tests and for including
GNU Prolog in his \MyUrl{http://logtalk.org}{logtalk} system (2000-).

Many thanks to John Collins, the
\MyUrl{https://ctan.org/pkg/latexmk/}{latexmk} maintainer, who greatly
helped me to simplify the building of the documentation using
\texttt{latexmk} (2021).


