\newpage
\section{Finite domain solver and built-in predicates}
%HEVEA\cutdef[1]{subsection}
\subsection{Introduction}
\label{Intro-FD}
The finite domain (FD) constraint solver extends Prolog with constraints over
FD. This facility is available if the FD part of GNU Prolog has been
installed.  The solver is an instance of the Constraint Logic Programming
scheme introduced by Jaffar and Lassez in 1987
\cite{Jaffar-Lassez87}. Constraints on FD are solved using propagation
techniques, in particular arc-consistency (AC). The interested reader can
refer to ``Constraint Satisfaction in Logic Programming'' of P. Van
Hentenryck (1989) \cite{pvh89}. The solver is based on the \texttt{clp(FD)}
solver \cite{long-clp-fd}. The GNU Prolog FD solver offers arithmetic
constraints, boolean constraints, reified constraints and symbolic
constraints on an new kind of variables: Finite Domain variables.

\subsubsection{Finite Domain variables}
\label{Finite-Domain-variables}
A new type of data is introduced: FD variables which can only take values in
their domains. The initial domain of an FD variable is
\texttt{0..fd\_max\_integer} where \IdxFKD{fd\_max\_integer} represents the
greatest value that any FD variable can take. The predicate
\texttt{fd\_max\_integer/1} returns this value which may be different from
the \IdxPF{max\_integer} \Idx{Prolog flag}
\RefSP{set-prolog-flag/2}. The domain of an FD variable \texttt{X} is
reduced step by step by constraints in a monotonic way: when a value
has been removed from the domain of \texttt{X} it will never reappear
in the domain of \texttt{X}. An FD variable is fully compatible with
both Prolog integers and Prolog variables.  Namely, when an FD
variable is expected by an FD constraint it is possible to pass a
Prolog integer (considered as an FD variable whose domain is a
singleton) or a Prolog variable (immediately bound to an initial range
\texttt{0..fd\_max\_integer}). This avoids the need for specific type
declaration. Although it is not necessary to declare the initial domain of an
FD variable (since it will be bound \texttt{0..fd\_max\_integer} when
appearing for the fist time in a constraint) it is advantageous to do so and
thus reduce as soon as possible the size of its domain: particularly because
GNU Prolog, for efficiency reasons, does not check for overflows. For instance,
without any preliminary domain definitions for \texttt{X}, \texttt{Y} and
\texttt{Z}, the non-linear constraint \texttt{X*Y\#=Z} will fail due to an
overflow when computing the upper bound of the domain of \texttt{Z}:
\texttt{fd\_max\_integer $\times$ fd\_max\_integer}. This overflow causes a
negative result for the upper bound and the constraint then fails.

There are two internal representations for an FD variable:

\begin{itemize}

\item \SPart{interval representation}: only the \emph{min} and the
  \emph{max} of the variable are maintained. In this representation it is
  possible to store values included in \texttt{0..fd\_max\_integer}.

\item \SPart{sparse representation}: an additional bit-vector is used to
  store the set of possible values for the variable (i.e. the domain). In
  this representation it is possible to store values included in
  \texttt{0..vector\_max}. By default \IdxFKD{vector\_max} is set to 127.
  This value can be redefined via an environment variable \texttt{VECTORMAX}
  or via the built-in predicate \IdxFB{fd\_set\_vector\_max/1}
  \RefSP{fd-set-vector-max/1}.  The predicate \IdxFB{fd\_vector\_max/1}
  returns the current value of \texttt{vector\_max}
  \RefSP{fd-max-integer/1}.

\end{itemize}

\index{extra-constrained|see {\texttt{extra\_cstr}}}
The initial representation for an FD variable \texttt{X} is always an
interval representation and is switched to a sparse representation when a
``hole'' appears in the domain (e.g. due to an inequality constraint). Once a
variable uses a sparse representation it will not switch back to an interval
representation even if there are no longer holes in its domain. When this
switching occurs some values in the domain of \texttt{X} can be lost since
\texttt{vector\_max} is less than \texttt{fd\_max\_integer}. We say that
``\texttt{X} is extra-constrained'' since
\texttt{X} is constrained by the solver to the domain
\texttt{0..vector\_max} (via an imaginary constraint
\texttt{X \#={\lt} \Param{vector\_max}}). An \IdxFKD{extra\_cstr} is
associated with each FD variable to indicate that values have been lost due to
the switch to a sparse representation. This flag is updated on every
operations. The domain of an extra-constrained FD variable is output followed
by the \texttt{@} symbol. When a constraint fails on a extra-constrained
variable a message \texttt{Warning: Vector too small - maybe lost solutions
  (FD Var:\Param{N})} is displayed (\Param{N} is the address of the involved
variable).

Example 1 (\texttt{vector\_max} = \texttt{127}):

\begin{tabular}{|l|l|c|l|}
\hline

Constraint on \texttt{X} & Domain of \texttt{X} & \texttt{extra\_cstr}
& Lost values \\

\hline\hline

\texttt{X \#={\lt} 512} & \texttt{0..512} & \texttt{off} & none \\

\hline

\texttt{X \#{\bs}= 10} & \texttt{0..9:11..127} & \texttt{on} &
\texttt{128..512} \\

\hline

\texttt{X \#={\lt} 100} & \texttt{0..9:11..100} & \texttt{off} & none \\

\hline
\end{tabular}

In this example, when the constraint \texttt{X \#{\bs}= 10} is posted some
values are lost, the \texttt{extra\_cstr} is then switched on. However,
posting the constraint \texttt{X \#={\lt} 100} will turn off the flag (no
values are lost).

Example 2:

\begin{tabular}{|l|l|c|l|}
\hline

Constraint on \texttt{X} & Domain of \texttt{X} & \texttt{extra\_cstr}
& Lost values \\

\hline

\texttt{X \#={\lt} 512} & \texttt{0..512} & \texttt{off} & none \\

\hline

\texttt{X \#{\bs}= 10} & \texttt{0..9:11..127} & \texttt{on} &
\texttt{128..512} \\

\hline

\texttt{X \#{\gt}= 256} & \texttt{Warning: Vector too small\ldots} &
\texttt{on} & \texttt{128..512} \\

\hline
\end{tabular}

In this example, the constraint \texttt{X \#{\gt}= 256} fails due to the lost
of \texttt{128..512} so a message is displayed onto the terminal. The
solution would consist in increasing the size of the vector either by setting
the environment variable \texttt{VECTORMAX} (e.g. to \texttt{512}) or using
\texttt{fd\_set\_vector\_max(512)}.

Finally, bit-vectors are not dynamic, i.e. all vectors have the same size
(\texttt{0..vector\_max}). So the use of \texttt{fd\_set\_vector\_max/1} is
limited to the initial definition of vector sizes and must occur before any
constraint. As seen before, the solver tries to display a message when a
failure occurs due to a too short \texttt{vector\_max}. Unfortunately, in
some cases it cannot detect the lost of values and no message is emitted. So
the user should always take care to this parameter to be sure that it is
large to encode any vector.

\subsection{FD variable parameters}

\subsubsection{\IdxFBD{fd\_max\_integer/1}\label{fd-max-integer/1}}

\begin{TemplatesOneCol}
fd\_max\_integer(?integer)

\end{TemplatesOneCol}

\Description

\texttt{fd\_max\_integer(N)} succeeds if \texttt{N} is the current value of
\IdxFK{fd\_max\_integer} \RefSP{Intro-FD}.

\begin{PlErrors}

\ErrCond{\texttt{N} is neither a variable nor an integer}
\ErrTerm{type\_error(integer, N)}

\end{PlErrors}

\Portability

GNU Prolog predicate.

\subsubsection{\IdxFBD{fd\_vector\_max/1}}

\begin{TemplatesOneCol}
fd\_vector\_max(?integer)

\end{TemplatesOneCol}

\Description

\texttt{fd\_vector\_max(N)} succeeds if \texttt{N} is the current value of
\IdxFK{vector\_max} \RefSP{Intro-FD}.

\begin{PlErrors}

\ErrCond{\texttt{N} is neither a variable nor an integer}
\ErrTerm{type\_error(integer, N)}

\end{PlErrors}

\Portability

GNU Prolog predicate.

\subsubsection{\IdxFBD{fd\_set\_vector\_max/1}\label{fd-set-vector-max/1}}

\begin{TemplatesOneCol}
fd\_set\_vector\_max(+integer)

\end{TemplatesOneCol}

\Description

\texttt{fd\_set\_vector\_max(N)} initializes \IdxFK{vector\_max} based on
the value \texttt{N} \RefSP{Intro-FD}. More precisely, on 32 bit
machines, \texttt{vector\_max} is set to the smallest value of
\texttt{(32*k)-}1 which is $\geq$ \texttt{N}.

\begin{PlErrors}

\ErrCond{\texttt{N} is a variable}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{N} is neither a variable nor an integer}
\ErrTerm{type\_error(integer, N)}

\ErrCond{\texttt{N} is an integer $<$ 0}
\ErrTerm{domain\_error(not\_less\_than\_zero, N)}

\end{PlErrors}

\Portability

GNU Prolog predicate.

\subsection{Initial value constraints}

\subsubsection{\IdxFBD{fd\_domain/3},
               \IdxFBD{fd\_domain\_bool/1}}

\begin{TemplatesOneCol}
fd\_domain(+fd\_variable\_list\_or\_fd\_variable, +integer,
+integer)\\
fd\_domain(?fd\_variable, +integer, +integer)\\
fd\_domain\_bool(+fd\_variable\_list)\\
fd\_domain\_bool(?fd\_variable)

\end{TemplatesOneCol}

\Description

\texttt{fd\_domain(Vars, Lower, Upper)} constraints each element \texttt{X}
of \texttt{Vars} to take a value in \texttt{Lower..Upper}. This
predicate is generally used to set the initial domain of variables to an
interval. \texttt{Vars} can be also a single FD variable (or a single Prolog
variable).

\texttt{fd\_domain\_bool(Vars)} is equivalent to \texttt{fd\_domain(Vars, 0,
1)} and is used to declare boolean FD variables.

\begin{PlErrors}

\ErrCond{\texttt{Vars} is not a variable but is a partial list}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{Vars} is neither a variable nor an FD variable nor an
integer nor a list}
\ErrTerm{type\_error(list, Vars)}

\ErrCond{an element \texttt{E} of the \texttt{Vars} list is neither a
variable nor an FD variable nor an integer}
\ErrTerm{type\_error(fd\_variable, E)}

\ErrCond{\texttt{Lower} is a variable}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{Lower} is neither a variable nor an integer}
\ErrTerm{type\_error(integer, Lower)}

\ErrCond{\texttt{Upper} is a variable}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{Upper} is neither a variable nor an integer}
\ErrTerm{type\_error(integer, Upper)}

\end{PlErrors}

\Portability

GNU Prolog predicate.

\subsubsection{\IdxFBD{fd\_domain/2}}

\begin{TemplatesOneCol}
fd\_domain(+fd\_variable\_list, +integer\_list)\\
fd\_domain(?fd\_variable, +integer\_list)

\end{TemplatesOneCol}

\Description

\texttt{fd\_domain(Vars, Values)} constraints each element \texttt{X} of the
list \texttt{Vars} to take a value in the list \texttt{Values}. This
predicate is generally used to set the initial domain of variables to a set
of values. The domain of each variable of \texttt{Vars} uses a sparse
representation. \texttt{Vars} can be also a single FD variable (or a single
Prolog variable).

\begin{PlErrors}

\ErrCond{\texttt{Vars} is not a variable but is a partial list}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{Vars} is neither a variable nor an FD variable nor an
integer nor a list}
\ErrTerm{type\_error(list, Vars)}

\ErrCond{an element \texttt{E} of the \texttt{Vars} list is neither a
variable nor an FD variable nor an integer}
\ErrTerm{type\_error(fd\_variable, E)}

\ErrCond{\texttt{Values} is a partial list or a list with an element
\texttt{E} which is a variable}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{Values} is neither a partial list nor a list}
\ErrTerm{type\_error(list, Values)}

\ErrCond{an element \texttt{E} of the \texttt{Values} list is neither a
variable nor an integer}
\ErrTerm{type\_error(integer, E)}

\end{PlErrors}

\Portability

GNU Prolog predicate.

\subsection{Type testing}

\subsubsection{\IdxFBD{fd\_var/1}, \IdxFBD{non\_fd\_var/1},
               \IdxFBD{generic\_var/1},
               \IdxFBD{non\_generic\_var/1}}

\begin{TemplatesTwoCols}
fd\_var(?term)\\
non\_fd\_var(?term)\\
generic\_var(?term)\\
non\_generic\_var(?term)

\end{TemplatesTwoCols}

\Description

\texttt{fd\_var(Term)} succeeds if \texttt{Term} is currently an
FD variable.

\texttt{non\_fd\_var(Term)} succeeds if \texttt{Term} is
currently not an FD variable (opposite of \texttt{fd\_var/1}).

\texttt{generic\_var(Term)} succeeds if \texttt{Term} is
either a Prolog variable or an FD variable.

\texttt{non\_generic\_var(Term)} succeeds if
\texttt{Term} is neither a Prolog variable nor an FD variable
(opposite of \texttt{generic\_var/1}).

\PlErrorsNone

\Portability

GNU Prolog predicate.

\subsection{FD variable information}
These predicate allow the user to get some information about FD variables.
They are not constraints, they only return the current state of a variable.

\subsubsection{\IdxFBD{fd\_min/2},
               \IdxFBD{fd\_max/2},
               \IdxFBD{fd\_size/2},
               \IdxFBD{fd\_dom/2}}

\begin{TemplatesOneCol}
fd\_min(+fd\_variable, ?integer)\\
fd\_max(+fd\_variable, ?integer)\\
fd\_size(+fd\_variable, ?integer)\\
fd\_dom(+fd\_variable, ?integer\_list)

\end{TemplatesOneCol}

\Description

\texttt{fd\_min(X, N)} succeeds if \texttt{N} is the minimal value of the
current domain of \texttt{X}.

\texttt{fd\_max(X, N)} succeeds if \texttt{N} is the maximal value of the
current domain of \texttt{X}.

\texttt{fd\_size(X, N)} succeeds if \texttt{N} is the number of elements of
the current domain of \texttt{X}.

\texttt{fd\_dom(X, Values)} succeeds if \texttt{Values} is the list of
values of the current domain of \texttt{X}.

\begin{PlErrors}

\ErrCond{\texttt{X} is a variable}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{X} is neither an FD variable nor an integer}
\ErrTerm{type\_error(fd\_variable, X)}

\ErrCond{\texttt{N} is neither a variable nor an integer}
\ErrTerm{type\_error(integer, N)}

\ErrCond{an element \texttt{E} of the \texttt{Vars} list is neither a
variable nor an FD variable nor an integer}
\ErrTerm{type\_error(fd\_variable, E)}

\ErrCond{\texttt{Values} is neither a partial list nor a list}
\ErrTerm{type\_error(list, Values)}

\end{PlErrors}

\Portability

GNU Prolog predicate.

\subsubsection{\IdxFBD{fd\_has\_extra\_cstr/1},
               \IdxFBD{fd\_has\_vector/1},
               \IdxFBD{fd\_use\_vector/1}}

\begin{TemplatesOneCol}
fd\_has\_extra\_cstr(+fd\_variable)\\
fd\_has\_vector(+fd\_variable)\\
fd\_use\_vector(+fd\_variable)

\end{TemplatesOneCol}

\Description

\texttt{fd\_has\_extra\_cstr(X)} succeeds if the \IdxFK{extra\_cstr}
of \texttt{X} is currently on \RefSP{Intro-FD}.

\texttt{fd\_has\_vector(X)} succeeds if the current domain of \texttt{X}
uses a sparse representation \RefSP{Intro-FD}.

\texttt{fd\_use\_vector(X)} enforces a sparse representation for the domain
of \texttt{X} \RefSP{Intro-FD}.

\begin{PlErrors}

\ErrCond{\texttt{X} is a variable}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{X} is neither an FD variable nor an integer}
\ErrTerm{type\_error(fd\_variable, X)}

\end{PlErrors}

\Portability

GNU Prolog predicates.

\subsection{Arithmetic constraints}

\subsubsection{FD arithmetic expressions}
\label{FD-arithmetic-expressions}
An FD arithmetic expression is a Prolog term built from integers, variables
(Prolog or FD variables), and functors (or operators) that represent
arithmetic functions. The following table details the components of an FD
arithmetic expression:


\begin{tabular}{|l|L{10cm}|}
\hline

FD Expression & Result \\

\hline\hline

Prolog variable & domain \texttt{0..fd\_max\_integer} \\

\hline

FD variable \texttt{X} & domain of \texttt{X} \\

\hline

integer number \texttt{N} & domain \texttt{N..N} \\

\hline

\texttt{+ E} & same as \texttt{E} \\

\hline

\texttt{- E} & opposite of \texttt{E} \\

\hline

\texttt{E1 + E2} & sum of \texttt{E1} and \texttt{E2} \\

\hline

\texttt{E1 - E2} & subtraction of \texttt{E2} from \texttt{E1} \\

\hline

\texttt{E1 * E2} & multiplication of \texttt{E1} by \texttt{E2} \\

\hline

\texttt{E1 / E2} & integer division of \texttt{E1} by \texttt{E2} (only
succeeds if the remainder is 0) \\

\hline

\texttt{E1 ** E2} & \texttt{E1} raised to the power of \texttt{E2
}(\texttt{E1} or \texttt{E2} must be an integer) \\

\hline

\texttt{min(E1,E2)} & minimum of \texttt{E1} and \texttt{E2} \\

\hline

\texttt{max(E1,E2)} & maximum of \texttt{E1} and \texttt{E2} \\

\hline

\texttt{dist(E1,E2)} & distance, i.e. $|$\texttt{E1 - E2$|$} \\

\hline

\texttt{E1 // E2} & quotient of the integer division of \texttt{E1} by
\texttt{E2} \\

\hline

\texttt{E1 rem E2} & remainder of the integer division of \texttt{E1} by
\texttt{E2} \\

\hline

\texttt{quot\_rem(E1,E2,R)} & quotient of the integer division of
\texttt{E1} by \texttt{E2}
  \linebreak
   (\texttt{R} is the remainder of the integer division of \texttt{E1} by
   \texttt{E2}) \\

\hline
\end{tabular}

FD expressions are not restricted to be linear. However non-linear
constraints usually yield less constraint propagation than linear
constraints.

\texttt{+}, \texttt{-}, \texttt{*}, \texttt{/}, \texttt{//}, \texttt{rem}
and \texttt{**} are predefined infix operators. \texttt{+} and \texttt{-}
are predefined prefix operators \RefSP{op/3:(Term-input/output)}.

\begin{PlErrors}

\ErrCond{a sub-expression is of the form \texttt{\_ ** E} and \texttt{E} is
a variable}
\ErrTerm{instantiation\_error}

\ErrCond{a sub-expression \texttt{E} is neither a variable nor an integer
nor an FD arithmetic functor}
\ErrTerm{type\_error(fd\_evaluable, E)}

\ErrCond{an expression is too complex}
\ErrTerm{resource\_error(too\_big\_fd\_constraint)}

\end{PlErrors}

\subsubsection{Partial AC: \IdxFBD{(\#=)/2} - constraint equal, \label{Partial-AC:-(:=)/2}
               \IdxFBD{(\#{\bs}=)/2} - constraint not equal, \\
               \IdxFBD{(\#{\lt})/2} - constraint less than,
               \IdxFBD{(\#={\lt})/2} - constraint less than or equal, \\
               \IdxFBD{(\#{\gt})/2} - constraint greater than,
               \IdxFBD{(\#{\gt}=)/2} - constraint greater than or equal}

\begin{TemplatesOneCol}
\#=(?fd\_evaluable, ?fd\_evaluable)\\
\#{\bs}=(?fd\_evaluable, ?fd\_evaluable)\\
\#{\lt}(?fd\_evaluable, ?fd\_evaluable)\\
\#={\lt}(?fd\_evaluable, ?fd\_evaluable)\\
\#{\gt}(?fd\_evaluable, ?fd\_evaluable)\\
\#{\gt}=(?fd\_evaluable, ?fd\_evaluable)

\end{TemplatesOneCol}

\Description

\texttt{FdExpr1 \#= FdExpr2} constrains \texttt{FdExpr1} to be
equal to \texttt{FdExpr2}.

\texttt{FdExpr1 \#{\bs}= FdExpr2} constrains \texttt{FdExpr1}
to be different from \texttt{FdExpr2}.

\texttt{FdExpr1 \#{\lt} FdExpr2} constrains \texttt{FdExpr1} to
be less than \texttt{FdExpr2}.

\texttt{FdExpr1 \#={\lt} FdExpr2} constrains \texttt{FdExpr1}
to be less than or equal to \texttt{FdExpr2}.

\texttt{FdExpr1 \#{\gt} FdExpr2} constrains \texttt{FdExpr1} to
be greater than \texttt{FdExpr2}.

\texttt{FdExpr1 \#{\gt}= FdExpr2} constrains \texttt{FdExpr1}
to be greater than or equal to \texttt{FdExpr2}.

\texttt{FdExpr1} and \texttt{FdExpr2} are arithmetic FD expressions
\RefSP{FD-arithmetic-expressions}.

\texttt{\#=}, \texttt{\#{\bs}=}, \texttt{\#{\lt}}, \texttt{\#={\lt}},
\texttt{\#{\gt}} and \texttt{\#{\gt}=} are predefined infix operators
\RefSP{op/3:(Term-input/output)}.

These predicates post arithmetic constraints that are managed by the solver
using a partial arc-consistency algorithm to reduce the domain of involved
variables. In this scheme only the bounds of the domain of variables are
updated. This leads to less propagation than full arc-consistency techniques
\RefSP{Full-AC:-(:=:)/2} but is generally more efficient for
arithmetic. These arithmetic constraints can be reified \RefSP{Boolean-and-reified-constraints}.

\Errors

Refer to the syntax of arithmetic FD expressions for possible errors
\RefSP{FD-arithmetic-expressions}.

\Portability

GNU Prolog predicates.

\subsubsection{Full AC: \IdxFBD{(\#=\#)/2} - constraint equal, \label{Full-AC:-(:=:)/2}
               \IdxFBD{(\#{\bs}=\#)/2} - constraint not equal, \\
               \IdxFBD{(\#{\lt}\#)/2} - constraint less than,
               \IdxFBD{(\#={\lt}\#)/2} - constraint less than or equal, \\
               \IdxFBD{(\#{\gt}\#)/2} - constraint greater than,
               \IdxFBD{(\#{\gt}=\#)/2} - constraint greater than or equal}

\begin{TemplatesOneCol}
\#=\#(?fd\_evaluable, ?fd\_evaluable)\\
\#{\bs}=\#(?fd\_evaluable, ?fd\_evaluable)\\
\#{\lt}\#(?fd\_evaluable, ?fd\_evaluable)\\
\#={\lt}\#(?fd\_evaluable, ?fd\_evaluable)\\
\#{\gt}\#(?fd\_evaluable, ?fd\_evaluable)\\
\#{\gt}=\#(?fd\_evaluable, ?fd\_evaluable)

\end{TemplatesOneCol}

\Description

\texttt{FdExpr1 \#=\# FdExpr2} constrains \texttt{FdExpr1} to
be equal to \texttt{FdExpr2}.

\texttt{FdExpr1 \#{\bs}=\# FdExpr2} constrains \texttt{FdExpr1} to be
different from \texttt{FdExpr2}.

\texttt{FdExpr1 \#{\lt}\# FdExpr2} constrains \texttt{FdExpr1}
to be less than \texttt{FdExpr2}.

\texttt{FdExpr1 \#={\lt}\# FdExpr2} constrains \texttt{FdExpr1} to be
less than or equal to \texttt{FdExpr2}.

\texttt{FdExpr1 \#{\gt}\# FdExpr2} constrains \texttt{FdExpr1}
to be greater than \texttt{FdExpr2}.

\texttt{FdExpr1 \#{\gt}=\# FdExpr2} constrains \texttt{FdExpr1} to be
greater than or equal to \texttt{FdExpr2}.

\texttt{FdExpr1} and \texttt{FdExpr2} are arithmetic FD expressions
\RefSP{FD-arithmetic-expressions}.

\texttt{\#=\#}, \texttt{\#{\bs}=\#}, \texttt{\#{\lt}\#},
\texttt{\#={\lt}\#}, \texttt{\#{\gt}\#} and \texttt{\#{\gt}=\#} are
predefined infix operators \RefSP{op/3:(Term-input/output)}.

These predicates post arithmetic constraints that are managed by the solver
using a full arc-consistency algorithm to reduce the domain of involved
variables. In this scheme the full domain of variables is updated. This
leads to more propagation than partial arc-consistency techniques \RefSP{FD-arithmetic-expressions} but is generally less efficient for arithmetic.
These arithmetic constraints can be reified \RefSP{Boolean-FD-expressions}.

\Errors

Refer to the syntax of arithmetic FD expressions for possible errors
\RefSP{FD-arithmetic-expressions}.

\Portability

GNU Prolog predicates.

\subsubsection{\IdxFBD{fd\_prime/1},
               \IdxFBD{fd\_not\_prime/1}}

\begin{TemplatesOneCol}
fd\_prime(?fd\_variable)\\
fd\_not\_prime(?fd\_variable)

\end{TemplatesOneCol}

\Description

\texttt{fd\_prime(X)} constraints \texttt{X} to be a prime number between
\texttt{0..\IdxFK{vector\_max}}.
This constraint enforces a sparse representation
for the domain of \texttt{X} \RefSP{Intro-FD}.

\texttt{fd\_not\_prime(X)} constraints \texttt{X} to be a non prime number
between \texttt{0..vector\_max}. This constraint enforces a sparse
representation for the domain of \texttt{X} \RefSP{Intro-FD}.

\begin{PlErrors}

\ErrCond{\texttt{X} is neither an FD variable nor an integer}
\ErrTerm{type\_error(fd\_variable, X)}

\end{PlErrors}

\Portability

GNU Prolog predicates.

\subsection{Boolean and reified constraints}
\label{Boolean-and-reified-constraints}

\subsubsection{Boolean FD expressions}
\label{Boolean-FD-expressions}

An boolean FD expression is a Prolog term built from integers (0 for false,
1 for true), variables (Prolog or FD variables), partial AC arithmetic
constraints \RefSP{Partial-AC:-(:=)/2}, full AC arithmetic constraints
\RefSP{Full-AC:-(:=:)/2} and functors (or operators) that represent
boolean functions. When a sub-expression of a boolean expression is an
arithmetic constraint \Param{c}, it is reified. Namely, as soon as the
solver detects that \Param{c} is true (i.e. \emph{entailed}) the
sub-expression has the value \texttt{1}. Similarly when the solver detects
that \Param{c} is false (i.e. \emph{disentailed}) the sub-expression
evaluates as \texttt{0}. While neither the entailment nor the disentailment
can be detected the sub-expression is evaluated as a domain \texttt{0..1}.
The following table details the components of an FD boolean expression:

\begin{tabular}{|l|l|}
\hline

FD Expression & Result \\

\hline\hline

Prolog variable & domain \texttt{0..1} \\

\hline

FD variable \texttt{X} & domain of \texttt{X}, \texttt{X} is constrained to
be in \texttt{0..1} \\

\hline

\texttt{0} (integer) & \texttt{0} (false) \\

\hline

\texttt{1} (integer) & \texttt{1} (true) \\

\hline

\texttt{\#{\bs} E} & not \texttt{E} \\

\hline

\texttt{E1 \#{\lt}={\gt} E2} & \texttt{E1} equivalent to \texttt{E2} \\

\hline

\texttt{E1 \#{\bs}{\lt}={\gt} E2} & \texttt{E1} not equivalent to
\texttt{E2} (i.e. \texttt{E1} different from \texttt{E2)} \\

\hline

\texttt{E1 \#\# E2} & \texttt{E1} exclusive OR \texttt{E2} (i.e. \texttt{E1}
not equivalent to \texttt{E2)} \\

\hline

\texttt{E1 \#=={\gt} E2} & \texttt{E1} implies \texttt{E2} \\

\hline

\texttt{E1 \#{\bs}=={\gt} E2} & \texttt{E1} does not imply \texttt{E2} \\

\hline

\texttt{E1 \#/{\bs} E2} & \texttt{E1} AND \texttt{E2} \\

\hline

\texttt{E1 \#{\bs}/{\bs} E2} & \texttt{E1} NAND \texttt{E2} \\

\hline

\texttt{E1 \#{\bs}/ E2} & \texttt{E1} OR \texttt{E2} \\

\hline

\texttt{E1 \#{\bs}{\bs}/ E2} & \texttt{E1} NOR \texttt{E2} \\

\hline
\end{tabular}

\texttt{\#{\lt}={\gt}}, \texttt{\#{\bs}{\lt}={\gt}}, \texttt{\#\#},
\texttt{\#=={\gt}}, \texttt{\#{\bs}=={\gt}}, \texttt{\#/{\bs}},
\texttt{\#{\bs}/{\bs}}, \texttt{\#{\bs}/} and \texttt{\#{\bs}{\bs}/} are
predefined infix operators. \texttt{\#{\bs}} is a predefined prefix operator
\RefSP{op/3:(Term-input/output)}.

\begin{PlErrors}

\ErrCond{a sub-expression \texttt{E} is neither a variable nor an integer (0
or 1) nor an FD boolean functor nor reified constraint}
\ErrTerm{type\_error(fd\_bool\_evaluable, E)}

\ErrCond{an expression is too complex}
\ErrTerm{resource\_error(too\_big\_fd\_constraint)}

\ErrCond{a sub-expression is an invalid reified constraint}
\ErrTermRm{an arithmetic constraint error \RefSP{FD-arithmetic-expressions}}

\end{PlErrors}



\subsubsection{\IdxFBD{fd\_reified\_in/4}}

\begin{TemplatesOneCol}
fd\_reified\_in(?fd\_variable, +integer, +integer, ?fd\_variable)
\end{TemplatesOneCol}

\Description

\texttt{fd\_reified\_in(X, Lower, Upper, B)} captures the truth value of the constraint $\texttt{X} \in [\texttt{Lower}..\texttt{Upper}]$ in the boolean variable \texttt{B}.

\begin{PlErrors}

\ErrCond{\texttt{X} is neither a variable nor an FD variable nor an integer}
\ErrTerm{type\_error(fd\_variable, X)}

\ErrCond{\texttt{B} is neither a variable nor an FD variable nor an integer}
\ErrTerm{type\_error(fd\_variable, B)}

\ErrCond{\texttt{Lower} is a variable}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{Lower} is neither a variable nor an integer}
\ErrTerm{type\_error(integer, Lower)}

\ErrCond{\texttt{Upper} is a variable}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{Upper} is neither a variable nor an integer}
\ErrTerm{type\_error(integer, Upper)}

\end{PlErrors}

\subsubsection{\IdxFBD{(\#{\bs})/1} - constraint NOT,
               \IdxFBD{(\#{\lt}={\gt})/2} - constraint equivalent, \\
               \IdxFBD{(\#{\bs}{\lt}={\gt})/2} - constraint different,
               \IdxFBD{(\#\#)/2} - constraint XOR, \\
               \IdxFBD{(\#=={\gt})/2} - constraint imply,
               \IdxFBD{(\#{\bs}=={\gt})/2} - constraint not imply, \\
               \IdxFBD{(\#/{\bs})/2} - constraint AND,
               \IdxFBD{(\#{\bs}/{\bs})/2} - constraint NAND, \\
               \IdxFBD{(\#{\bs}/)/2} - constraint OR,
               \IdxFBD{(\#{\bs}{\bs}/)/2} - constraint NOR}

\begin{TemplatesOneCol}
\#{\bs}(?fd\_bool\_evaluable)\\
\#{\lt}={\gt}(?fd\_bool\_evaluable, ?fd\_bool\_evaluable)\\
\#{\bs}{\lt}={\gt}(?fd\_bool\_evaluable, ?fd\_bool\_evaluable)\\
\#\#(?fd\_bool\_evaluable, ?fd\_bool\_evaluable)\\
\#=={\gt}(?fd\_bool\_evaluable, ?fd\_bool\_evaluable)\\
\#{\bs}=={\gt}(?fd\_bool\_evaluable, ?fd\_bool\_evaluable)\\
\#/{\bs}(?fd\_bool\_evaluable, ?fd\_bool\_evaluable)\\
\#{\bs}/{\bs}(?fd\_bool\_evaluable, ?fd\_bool\_evaluable)\\
\#{\bs}/(?fd\_bool\_evaluable, ?fd\_bool\_evaluable)\\
\#{\bs}{\bs}/(?fd\_bool\_evaluable, ?fd\_bool\_evaluable)

\end{TemplatesOneCol}

\Description

\texttt{\#{\bs}= FdBoolExpr1} constraints \texttt{FdBoolExpr1} to be
false.

\texttt{FdBoolExpr1 \#{\lt}={\gt} FdBoolExpr2} constrains
\texttt{FdBoolExpr1} to be equivalent to \texttt{FdBoolExpr2}.

\texttt{FdBoolExpr1 \#{\bs}{\lt}={\gt} FdBoolExpr2} constrains
\texttt{FdBoolExpr1} to be equivalent to not \texttt{FdBoolExpr2}.

\texttt{FdBoolExpr1 \#\# FdBoolExpr2} constrains \texttt{FdBoolExpr1} XOR
\texttt{FdBoolExpr2} to be true

\texttt{FdBoolExpr1 \#=={\gt} FdBoolExpr2} constrains
\texttt{FdBoolExpr1} to imply \texttt{FdBoolExpr2}.

\texttt{FdBoolExpr1 \#{\bs}=={\gt} FdBoolExpr2} constrains
\texttt{FdBoolExpr1} to not imply \texttt{FdBoolExpr2}.

\texttt{FdBoolExpr1 \#/{\bs} FdBoolExpr2} constrains \texttt{FdBoolExpr1}
AND \texttt{FdBoolExpr2} to be true.

\texttt{FdBoolExpr1 \#{\bs}/{\bs} FdBoolExpr2} constrains
\texttt{FdBoolExpr1} AND \texttt{FdBoolExpr2} to be false.

\texttt{FdBoolExpr1 \#{\bs}/ FdBoolExpr2} constrains \texttt{FdBoolExpr1}
OR \texttt{FdBoolExpr2} to be true.

\texttt{FdBoolExpr1 \#{\bs}{\bs}/ FdBoolExpr2} constrains
\texttt{FdBoolExpr1} OR \texttt{FdBoolExpr2} to be false.

\texttt{FdBoolExpr1} and \texttt{FdBoolExpr2} are boolean FD expressions
\RefSP{Boolean-FD-expressions}.

Note that \texttt{\#{\bs}{\lt}={\gt}} (not equivalent) and \texttt{\#\#}
(exclusive or) are synonymous.

These predicates post boolean constraints that are managed by the FD solver
using a partial arc-consistency algorithm to reduce the domain of involved
variables. The (dis)entailment of reified constraints is detected using
either the bounds (for partial AC arithmetic constraints) or the full domain
(for full AC arithmetic constraints).

\texttt{\#{\lt}={\gt}}, \texttt{\#{\bs}{\lt}={\gt}}, \texttt{\#\#},
\texttt{\#=={\gt}}, \texttt{\#{\bs}=={\gt}}, \texttt{\#/{\bs}},
\texttt{\#{\bs}/{\bs}}, \texttt{\#{\bs}/} and \texttt{\#{\bs}{\bs}/} are
predefined infix operators. \texttt{\#{\bs}} is a predefined prefix operator
\RefSP{op/3:(Term-input/output)}.

\Errors

Refer to the syntax of boolean FD expressions for possible errors
\RefSP{Boolean-FD-expressions}.

\Portability

GNU Prolog predicates.

\subsubsection{\IdxFBD{fd\_cardinality/2},\label{fd-cardinality/2}
               \IdxFBD{fd\_cardinality/3},
               \IdxFBD{fd\_at\_least\_one/1},
               \IdxFBD{fd\_at\_most\_one/1}, \\
               \IdxFBD{fd\_only\_one/1}}

\begin{TemplatesOneCol}
fd\_cardinality(+fd\_bool\_evaluable\_list, ?fd\_variable)\\
fd\_cardinality(+integer, ?fd\_variable, +integer)\\
fd\_at\_least\_one(+fd\_bool\_evaluable\_list)\\
fd\_at\_most\_one(+fd\_bool\_evaluable\_list)\\
fd\_only\_one(+fd\_bool\_evaluable\_list)

\end{TemplatesOneCol}

\Description

\texttt{fd\_cardinality(List, Count)} unifies \texttt{Count} with the number
of constraints that are true in \texttt{List}. This is equivalent to post
the constraint \texttt{B$_{1}$ + B$_{2}$ + \ldots + B$_{n}$ \#= Count}
where each variable \texttt{Bi} is a new variable defined by the constraint
\texttt{B$_{i}$ \#{\lt}={\gt} C$_{i}$} where \texttt{C$_{i}$} is the
\texttt{i}\emph{th} constraint of \texttt{List}. Each \texttt{C$_{i}$}
must be a boolean FD expression \RefSP{Boolean-FD-expressions}.

\texttt{fd\_cardinality(Lower, List, Upper)} is equivalent to
\texttt{fd\_cardinality(List, Count), Lower \#={\lt} Count, Count \#={\lt}
Upper}

\texttt{fd\_at\_least\_one(List)} is equivalent to
\texttt{fd\_cardinality(List, Count), Count \#{\gt}= 1}.

\texttt{fd\_at\_most\_one(List)} is equivalent to
\texttt{fd\_cardinality(List, Count), Count \#={\lt} 1}.

\texttt{fd\_only\_one(List)} is equivalent to \texttt{fd\_cardinality(List,
1)}.

\begin{PlErrors}

\ErrCond{\texttt{List} is a partial list}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{List} is neither a partial list nor a list}
\ErrTerm{type\_error(list, List)}

\ErrCond{\texttt{Count} is neither an FD variable nor an integer}
\ErrTerm{type\_error(fd\_variable, Count)}

\ErrCond{\texttt{Lower} is a variable}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{Lower} is neither a variable nor an integer}
\ErrTerm{type\_error(integer, Lower)}

\ErrCond{\texttt{Upper} is a variable}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{Upper} is neither a variable nor an integer}
\ErrTerm{type\_error(integer, Upper)}

\ErrCond{an element \texttt{E} of the \texttt{List} list is an invalid
boolean expression}
\ErrTermRm{an FD boolean constraint \RefSP{Boolean-FD-expressions}}

\end{PlErrors}

\Portability

GNU Prolog predicates.

\subsection{Symbolic constraints}

\subsubsection{\IdxFBD{fd\_all\_different/1}}

\begin{TemplatesOneCol}
fd\_all\_different(+fd\_variable\_list)

\end{TemplatesOneCol}

\Description

\texttt{fd\_all\_different(List)} constrains all variables in \texttt{List}
to take distinct values. This is equivalent to posting an inequality
constraint for each pair of variables. This constraint is triggered when a
variable becomes ground, removing its value from the domain of the other
variables.

\begin{PlErrors}

\ErrCond{\texttt{List} is a partial list}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{List} is neither a partial list nor a list}
\ErrTerm{type\_error(list, List)}

\ErrCond{an element \texttt{E} of the \texttt{List} list is neither a
variable nor an integer nor an FD variable}
\ErrTerm{type\_error(fd\_variable, E)}

\end{PlErrors}

\Portability

GNU Prolog predicate.

\subsubsection{\IdxFBD{fd\_element/3}\label{fd-element/3}}

\begin{TemplatesOneCol}
fd\_element(?fd\_variable, +integer\_list, ?fd\_variable)

\end{TemplatesOneCol}

\Description

\texttt{fd\_element(I, List, X)} constraints \texttt{X} to be equal to the
\texttt{I}\emph{th} integer (from 1) of \texttt{List}.

\begin{PlErrors}

\ErrCond{\texttt{I} is neither a variable nor an FD variable nor an integer}
\ErrTerm{type\_error(fd\_variable, I)}

\ErrCond{\texttt{X} is neither a variable nor an FD variable nor an integer}
\ErrTerm{type\_error(fd\_variable, X)}

\ErrCond{\texttt{List} is a partial list or a list with an element
\texttt{E} which is a variable}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{List} is neither a partial list nor a list}
\ErrTerm{type\_error(list, List)}

\ErrCond{an element \texttt{E} of the \texttt{List} list is neither a
variable nor an integer}
\ErrTerm{type\_error(integer, E)}

\end{PlErrors}

\Portability

GNU Prolog predicate.

\subsubsection{\IdxFBD{fd\_element\_var/3}}

\begin{TemplatesOneCol}
fd\_element\_var(?fd\_variable, +fd\_variable\_list, ?fd\_variable)

\end{TemplatesOneCol}

\Description

\texttt{fd\_element\_var(I, List, X)} constraints \texttt{X} to be equal to
the \texttt{I}\emph{th} variable (from 1) of \texttt{List}. This
constraint is similar to \texttt{fd\_element/3} \RefSP{fd-element/3} but
\texttt{List} can also contain FD variables (rather than just integers).

\begin{PlErrors}

\ErrCond{\texttt{I} is neither a variable nor an FD variable nor an integer}
\ErrTerm{type\_error(fd\_variable, I)}

\ErrCond{\texttt{X} is neither a variable nor an FD variable nor an integer}
\ErrTerm{type\_error(fd\_variable, X)}

\ErrCond{\texttt{List} is a partial list}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{List} is neither a partial list nor a list}
\ErrTerm{type\_error(list, List)}

\ErrCond{an element \texttt{E} of the \texttt{List} list is neither a
variable nor an integer nor an FD variable}
\ErrTerm{type\_error(fd\_variable, E)}

\end{PlErrors}

\Portability

GNU Prolog predicate.

\subsubsection{\IdxFBD{fd\_atmost/3},
               \IdxFBD{fd\_atleast/3},
               \IdxFBD{fd\_exactly/3}}

\begin{TemplatesOneCol}
fd\_atmost(+integer, +fd\_variable\_list, +integer)\\
fd\_atleast(+integer, +fd\_variable\_list, +integer)\\
fd\_exactly(+integer, +fd\_variable\_list, +integer)

\end{TemplatesOneCol}

\Description

\texttt{fd\_atmost(N, List, V)} posts the constraint that at most \texttt{N}
variables of \texttt{List} are equal to the value \texttt{V}.

\texttt{fd\_atleast(N, List, V)} posts the constraint that at least
\texttt{N} variables of \texttt{List} are equal to the value \texttt{V}.

\texttt{fd\_exactly(N, List, V)} posts the constraint that at exactly
\texttt{N} variables of \texttt{List} are equal to the value \texttt{V}.

These constraints are special cases of \IdxFB{fd\_cardinality/2}
\RefSP{fd-cardinality/2} but their implementation is more efficient.

\begin{PlErrors}

\ErrCond{\texttt{N} is a variable}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{N} is neither a variable nor an integer}
\ErrTerm{type\_error(integer, N)}

\ErrCond{\texttt{V} is a variable}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{V} is neither a variable nor an integer}
\ErrTerm{type\_error(integer, V)}

\ErrCond{\texttt{List} is a partial list}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{List} is neither a partial list nor a list}
\ErrTerm{type\_error(list, List)}

\ErrCond{an element \texttt{E} of the \texttt{List} list is neither a
variable nor an FD variable nor an integer}
\ErrTerm{type\_error(fd\_variable, E)}

\end{PlErrors}

\Portability

GNU Prolog predicates.

\subsubsection{\IdxFBD{fd\_relation/2},
               \IdxFBD{fd\_relationc/2}}

\begin{TemplatesOneCol}
fd\_relation(+integer\_list\_list, ?fd\_variable\_list)\\
fd\_relationc(+integer\_list\_list, ?fd\_variable\_list)

\end{TemplatesOneCol}

\Description

\texttt{fd\_relation(Relation, Vars)} constraints the tuple of variables
\texttt{Vars} to be equal to one tuple of the list \texttt{Relation}. A
tuple is represented by a list.

Example: definition of the boolean AND relation so that X AND Y
$\Leftrightarrow$ Z:

\begin{Indentation}
\begin{verbatim}
and(X,Y,Z):-
        fd_relation([[0,0,0],[0,1,0],[1,0,0],[1,1,1]], [X,Y,Z]).
\end{verbatim}
\end{Indentation}

\texttt{fd\_relationc(Columns, Vars)} is similar to \texttt{fd\_relation/2}
except that the relation is not given as the list of tuples but as the list
of the columns of the relation. A column is represented by a list.

Example:

\begin{Indentation}
\begin{verbatim}
and(X,Y,Z):-
        fd_relationc([[0,0,1,1],[0,1,0,1],[0,0,0,1]], [X,Y,Z]).
\end{verbatim}
\end{Indentation}

\begin{PlErrors}

\ErrCond{\texttt{Relation} is a partial list or a list with a sub-term
\texttt{E} which is a variable}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{Relation} is neither a partial list nor a list}
\ErrTerm{type\_error(list, Relation)}

\ErrCond{an element \texttt{E} of the \texttt{Relation} list is neither a
variable nor an integer}
\ErrTerm{type\_error(integer, E)}

\ErrCond{\texttt{Vars} is a partial list}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{Vars} is neither a partial list nor a list}
\ErrTerm{type\_error(list, Vars)}

\ErrCond{an element \texttt{E} of the \texttt{Vars} list is neither a
variable nor an integer nor an FD variable}
\ErrTerm{type\_error(fd\_variable, E)}

\end{PlErrors}

\Portability

GNU Prolog predicates.

\subsection{Labeling constraints}

\subsubsection{\IdxFBD{fd\_labeling/2},\label{fd-labeling/2}
               \IdxFBD{fd\_labeling/1},
               \IdxFBD{fd\_labelingff/1}}

\begin{TemplatesOneCol}
fd\_labeling(+fd\_variable\_list, +fd\_labeling\_option\_list)\\
fd\_labeling(+fd\_variable, +fd\_labeling\_option\_list)\\
fd\_labeling(+fd\_variable\_list)\\
fd\_labeling(+fd\_variable)\\
fd\_labelingff(+fd\_variable\_list)\\
fd\_labelingff(+fd\_variable)

\end{TemplatesOneCol}

\Description

\texttt{fd\_labeling(Vars, Options)} assigns a value to each variable
\texttt{X} of the list \texttt{Vars} according to the list of labeling
options given by \texttt{Options}. \texttt{Vars} can be also a single FD
variable. This predicate is re-executable on backtracking.

\SPart{FD labeling options}: \texttt{Options} is a list of labeling
options. If this list contains contradictory options, the rightmost option
is the one which applies. Possible options are:

\begin{itemize}

\item \AddFOD{variable\_method}\texttt{variable\_method(V)}: specifies the
heuristics to select the variable to enumerate:

\begin{itemize}

\item \IdxFOD{standard}: no heuristics, the leftmost variable is selected.

\item \IdxFOD{first\_fail} (or \texttt{ff}): selects the variable with the
smallest number of elements in its domain. If several variables have the
same number of elements the leftmost variable is selected.

\item \IdxFOD{most\_constrained}: like \texttt{first\_fail} but when
several variables have the same number of elements selects the
variable that appears in most constraints.

\item \IdxFOD{smallest}: selects the variable that has the smallest value
in its domain. If there is more than one such variable selects the
variable that appears in most constraints.

\item \IdxFOD{largest}: selects the variable that has the greatest value in
its domain. If there is more than one such variable selects the variable
that appears in most constraints.

\item \IdxFOD{max\_regret}: selects the variable that has the greatest
difference between the smallest value and the next value of its domain. If
there is more than one such variable selects the variable that appears in
most constraints.

\item \IdxFOD{random}: selects randomly a variable. Each variable is 
chosen only once.

\end{itemize}

\BL The default value is \texttt{standard}.

\item \AddFOD{reorder}\texttt{reorder(true/false)}: specifies if the variable
heuristics should dynamically reorder the list of variable (\texttt{true}) or
not (\texttt{false}). Dynamic reordering is generally more efficient but in
some cases a static ordering is faster. The default value is
\texttt{true}.

\item \AddFOD{value\_method}\texttt{value\_method(V)}: specifies the heuristics
to select the value to assign to the chosen variable:

\begin{itemize}

\item \IdxFOD{min}: enumerates the values from the smallest to the greatest
(default).

\item \IdxFOD{max}: enumerates the values from the greatest to the smallest.

\item \IdxFOD{middle}: enumerates the values from the middle to the bounds.

\item \IdxFOD{bounds}: enumerates the values from the bounds to the middle.

\item \IdxFOD{random}: enumerates the values randomly. Each value is tried
  only once.

\item \IdxFOD{bisect}: recursively creates a choice between \texttt{X \#={\lt}
    M} and \texttt{X \#{\gt} M}, where \texttt{M} is the midpoint of the
  domain of the variable. Values are thus tried from the smallest to the
  greatest. This is also known as \textit{domain splitting}.

\end{itemize}

\BL The default value is \texttt{min}.

\item \AddFOD{backtracks}\texttt{backtracks(B)}: unifies \texttt{B} with the
number of backtracks during the enumeration.

\end{itemize}

\texttt{fd\_labeling(Vars)} is equivalent to \texttt{fd\_labeling(Vars,
[])}.

\texttt{fd\_labelingff(Vars)} is equivalent to \texttt{fd\_labeling(Vars,
[variable\_method(ff)])}.

\begin{PlErrors}

\ErrCond{\texttt{Vars} is a partial list or a list with an element
\texttt{E} which is a variable}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{Vars} is neither a partial list nor a list}
\ErrTerm{type\_error(list, Vars)}

\ErrCond{an element \texttt{E} of the \texttt{Vars} list is neither a
variable nor an integer nor an FD variable}
\ErrTerm{type\_error(fd\_variable, E)}

\ErrCond{\texttt{Options} is a partial list or a list with an element
\texttt{E} which is a variable}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{Options} is neither a partial list nor a list}
\ErrTerm{type\_error(list, Options)}

\ErrCond{an element \texttt{E} of the \texttt{Options} list is neither a
variable nor a labeling option}
\ErrTerm{domain\_error(fd\_labeling\_option, E)}

\end{PlErrors}

\Portability

GNU Prolog predicates.

\subsection{Optimization constraints}

\subsubsection{\IdxFBD{fd\_minimize/2},
               \IdxFBD{fd\_maximize/2}}

\begin{TemplatesOneCol}
fd\_minimize(+callable\_term, ?fd\_variable)\\
fd\_maximize(+callable\_term, ?fd\_variable)

\end{TemplatesOneCol}

\Description

\texttt{fd\_minimize(Goal, X)} repeatedly calls \texttt{Goal} to find a
value that minimizes the variable \texttt{X}. \texttt{Goal} is a Prolog goal
that should instantiate \texttt{X}, a common case being the use of
\IdxFB{fd\_labeling/2} \RefSP{fd-labeling/2}. This predicate uses a
branch-and-bound algorithm with restart: each time \texttt{call(Goal)}
succeeds the computation restarts with a new constraint \texttt{X \#{\lt} V}
where \texttt{V} is the value of \texttt{X} at the end of the last call of
\texttt{Goal}. When a failure occurs (either because there are no remaining
choice-points for \texttt{Goal} or because the added constraint is
inconsistent with the rest of the store) the last solution is recomputed
since it is optimal.

\texttt{fd\_maximize(Goal, X)} is similar to \texttt{fd\_minimize/2} but
\texttt{X} is maximized\texttt{.}

\begin{PlErrors}

\ErrCond{\texttt{Goal} is a variable}
\ErrTerm{instantiation\_error}

\ErrCond{\texttt{Goal} is neither a variable nor a callable term}
\ErrTerm{type\_error(callable, Goal)}

\ErrCond{The predicate indicator \texttt{Pred} of \texttt{Goal} does not
correspond to an existing procedure and the value of the \texttt{unknown}
Prolog flag is \texttt{error} \RefSP{set-prolog-flag/2}}
\ErrTerm{existence\_error(procedure, Pred)}

\ErrCond{\texttt{X} is neither a variable nor an FD variable nor an integer
}
\ErrTerm{type\_error(fd\_variable, X)}

\end{PlErrors}

\Portability

GNU Prolog predicates.
%HEVEA\cutend
