\newpage
%BEGIN LATEX
\addcontentsline{toc}{section}{\numberline{}References}
%END LATEX


\begin{thebibliography}{99}

\bibitem{Ait-Kaci91} H. A\"{\i}t-Kaci.
\newblock ``Warren's Abstract Machine, A Tutorial Reconstruction''. \\
\newblock Logic Programming Series, MIT Press, 1991. \\
\newblock \OneUrl{http://web.archive.org/web/20071225092145/www.vanx.org/archive/wam/wam.html}

\bibitem{Clock} W.F. Clocksin and C.S. Mellish.
\newblock Programming in Prolog, Springer-Verlag, 1981.

\bibitem{wamcc} P. Codognet and D. Diaz.
\newblock ``{\tt wamcc}: Compiling Prolog to C''.\\
\newblock In {\em 12th International Conference on Logic Programming},
Tokyo, Japan, MIT Press, 1995. \\
\newblock \OneUrl{http://cri-dist.univ-paris1.fr/diaz/publications/WAMCC/iclp95.pdf}


\bibitem{long-clp-fd} P. Codognet and D. Diaz.
\newblock ``Compiling Constraint in {\tt clp(FD)}''. \\
\newblock {\em Journal of Logic Programming}, Vol. 27, No. 3, June 1996. \\
\newblock \OneUrl{http://cri-dist.univ-paris1.fr/diaz/publications/CLP-FD/jlp96.pdf}

\bibitem{gnu-prolog} D. Diaz and P. Codognet.
\newblock ``Design and Implementation of the GNU Prolog System''. \\
\newblock {\em Journal of Functional and Logic Programming}, Vol. 2001, No. 6, October 2001. \\
\newblock \OneUrl{http://cri-dist.univ-paris1.fr/diaz/publications/GNU-PROLOG/jflp01.pdf}

\bibitem{iso-part1}
\newblock Information technology - Programming languages - Prolog - Part 1:
General Core. \\
\newblock ISO/IEC 13211-1, 1995.
%\newblock \OneUrl{http://www.logic-programming.org/prolog\_std.html}

\bibitem{Jaffar-Lassez87} J. Jaffar and J-L. Lassez.
\newblock ``Constraint Logic Programming''. \\
\newblock In {\em Principles Of Programming Languages},
Munich, Germany, January 1987.

\bibitem{pvh89} P. Van Hentenryck.
\newblock ``Constraint Satisfaction in Logic Programming''. \\
\newblock Logic Programming Series, The MIT Press, 1989.

\bibitem{Warren83} D. H. D. Warren.
\newblock ``An Abstract Prolog Instruction Set''. \\
\newblock Technical Report 309, SRI International, Oct. 1983.

\end{thebibliography}
